---
title: "covid_psqi_t1"
author: "Jackie"
date: "7/4/2020"
output: html_notebook
---

# loading libraries
```{r}
library(lmerTest)
library(tidyverse)
library(foreign)
library(haven)
library(readxl)
library(naniar)
library(lubridate)
```

# reading in data
```{r}
covidt1_fp <- "~/ELS_COVID_T1_Final_Child_Raw.csv"
covidt1 <- read_csv(covidt1_fp)
```

## selecting relevant variables
```{r}
covidt1_select <-
  covidt1 %>%
  dplyr::select(ELS_ID, Child_Age.TC, Hours_sleep_3months.TC, contains("psqi"))
```

_Only include self-rated questions, which yield seven components of sleep quality:_ 0 = no difficulty, 3 = severe difficulty  
No reverse scoring  
The global score can range from 0 - 21

### _1_ Subjective Sleep Quality  
```{r}
# question 9 in the way that we ordered it ("During the past month, how would you rate your sleep
# quality over all?")
covidt1_comp1 <- 
  covidt1_select %>%
  mutate(comp1_subj_sleep_qual = covid_psqi_9)
```

### _2_ Sleep latency

#### _step 1 wrangling_
```{r}
# if people wrote a range of minutes for sleep latency: separated the min and max into two spearate variables

covidt1_comp2 <- 
  covidt1_comp1 %>%
  mutate(
    covid_psqi_2_copy = covid_psqi_2
    ) %>%
  mutate(
    covid_psqi_2_copy = str_replace(covid_psqi_2_copy, "May", "5"), # some answered "5" but exported as "May"
    covid_psqi_2_clean = str_extract(covid_psqi_2_copy, "\\-*\\d+\\.*\\d*"),
    covid_psqi_2_pt1 = ifelse(str_detect(covid_psqi_2_copy, "-"),
                              str_extract(covid_psqi_2_copy, "\\d+"),
                              NA), # if there was a range extracting the min and max
    covid_psqi_2_pt2 = ifelse(str_detect(covid_psqi_2_copy, "-"),
                              str_extract(covid_psqi_2_copy, "\\-\\d+\\.*"),
                              NA),
    covid_psqi_2_pt2 = str_extract(covid_psqi_2_pt2, "\\d+")
  ) 


# dropping incomplete psqis
covidt1_comp2_clean <-
  covidt1_comp2 %>%
  mutate(
    covid_psqi_2_pt1 = as.numeric(as.character(covid_psqi_2_pt1)),
    covid_psqi_2_pt2 = as.numeric(as.character(covid_psqi_2_pt2))
  ) %>%
  drop_na(psqi_timestamp)
```

#### _step 2_ averaging min and max of range for q2
```{r}
# averaging min and max, otherwise using the clean variable
covidt1_comp2_clean <-
  covidt1_comp2_clean %>%
  tibble::as_tibble() %>%
  group_by(ELS_ID) %>%
  mutate(
    covid_psqi_2_clean_rec =
      ifelse(!is.na(covid_psqi_2_pt1) &
               !is.na(covid_psqi_2_pt2),
             (covid_psqi_2_pt1 + covid_psqi_2_pt2)/2,
             as.numeric(covid_psqi_2_clean)
             )
    )
```

#### recoding q2, adding q2 and q5a, and recoding for final component 2 score
```{r}
covidt1_comp2_clean_sum <-
  covidt1_comp2_clean %>%
  dplyr::select(-covid_psqi_2_copy,
                -covid_psqi_2_clean,
                -covid_psqi_2_pt1,
                -covid_psqi_2_pt2) %>%
  mutate(
    comp2_psqi_2_clean_rec_fin =
      ifelse(covid_psqi_2_clean_rec <= 15,
             "0",
             "1"),
    comp2_psqi_2_clean_rec_fin = 
      ifelse(covid_psqi_2_clean_rec >= 16 & covid_psqi_2_clean_rec <= 30,
             "1",
             comp2_psqi_2_clean_rec_fin),
    comp2_psqi_2_clean_rec_fin =
      ifelse(covid_psqi_2_clean_rec >= 31 & covid_psqi_2_clean_rec <= 60,
             "2",
             comp2_psqi_2_clean_rec_fin),
    comp2_psqi_2_clean_rec_fin =
      ifelse(covid_psqi_2_clean_rec > 60,
             "3",
             comp2_psqi_2_clean_rec_fin)
    ) %>%
  mutate(
    comp2_psqi_2_clean_rec_fin = as.numeric(comp2_psqi_2_clean_rec_fin),
    covid_psqi_5a = as.numeric(covid_psqi_5a),
    comp2_psqi_2plus5a =
      (comp2_psqi_2_clean_rec_fin + covid_psqi_5a) # adding q2 and q5a
  ) %>%
  mutate(
    comp2_sleep_lat =
      ifelse(comp2_psqi_2plus5a == 0,
             "0",
             "1"),
    comp2_sleep_lat =
      ifelse(comp2_psqi_2plus5a >= 1 & comp2_psqi_2plus5a <= 2,
             "1",
             comp2_sleep_lat),
    comp2_sleep_lat =
      ifelse(comp2_psqi_2plus5a >=3 & comp2_psqi_2plus5a <= 4,
             "2",
             comp2_sleep_lat),
    comp2_sleep_lat =
      ifelse(comp2_psqi_2plus5a >=5 & comp2_psqi_2plus5a <= 6,
             "3",
             comp2_sleep_lat)
  )

covidt1_comp2_clean_sum <-
  covidt1_comp2_clean_sum %>%
  dplyr::select(-covid_psqi_2_clean_rec,
                -comp2_psqi_2_clean_rec_fin,
                -comp2_psqi_2plus5a
                )
```

### _3_ Sleep Duration

#### _step 1 wrangling_
```{r}
covidt1_comp3 <- 
  covidt1_comp2_clean_sum %>%
  mutate(
    covid_psqi_4_copy = covid_psqi_4
    ) %>%
  mutate(
    covid_psqi_4_copy = str_replace(covid_psqi_4_copy, "May", "5"),# some answered "5" but exported as "May"
    covid_psqi_4_copy = str_replace(covid_psqi_4_copy, "Jun", "6"), # some answered "6" but exported as "Jun"
    covid_psqi_4_copy = str_replace(covid_psqi_4_copy, "Jul", "7"), # some answered "7" but exported as "Jul"
    covid_psqi_4_copy = str_replace(covid_psqi_4_copy, "Aug", "8"), # some answered "8" but exported as "Aug"
    covid_psqi_4_clean = str_replace(covid_psqi_4_copy, ":", "."),
    covid_psqi_4_clean = str_extract(covid_psqi_4_clean, "\\.*\\d+\\.*\\d*"),
    covid_psqi_4_pt1 = ifelse(str_detect(covid_psqi_4_copy, "-"),
                              str_extract(covid_psqi_4_copy, "\\d+"),
                              NA), # if there was a range extracting the min and max
    covid_psqi_4_pt2 = ifelse(str_detect(covid_psqi_4_copy, "-"),
                              str_extract(covid_psqi_4_copy, "\\-\\d+\\.*"),
                              NA),
    covid_psqi_4_pt2 = str_extract(covid_psqi_4_pt2, "\\d+")
  )
  
# converting to numeric
covidt1_comp3_clean <-
  covidt1_comp3 %>%
  mutate(
    covid_psqi_4_pt1 = as.numeric(as.character(covid_psqi_4_pt1)),
    covid_psqi_4_pt2 = as.numeric(as.character(covid_psqi_4_pt2))
  )
  
```

#### _step 2_ averaging min and max of range for q4
```{r}
# averaging min and max, otherwise using the clean variable
covidt1_comp3_clean <-
  covidt1_comp3_clean %>%
  tibble::as_tibble() %>%
  group_by(ELS_ID) %>%
  mutate(
    covid_psqi_4_clean_rec =
      ifelse(!is.na(covid_psqi_4_pt1) &
               !is.na(covid_psqi_4_pt2),
             (covid_psqi_4_pt1 + covid_psqi_4_pt2)/2,
             as.numeric(covid_psqi_4_clean)
             )
    )

covidt1_comp3_clean_sum <-
  covidt1_comp3_clean %>%
  mutate(
    comp3_sleep_dur =
      ifelse(covid_psqi_4_clean_rec >7,
             "0",
             "1"),
    comp3_sleep_dur = 
      ifelse(covid_psqi_4_clean_rec >= 6 & covid_psqi_4_clean_rec <= 7,
             "1",
             comp3_sleep_dur),
    comp3_sleep_dur =
      ifelse(covid_psqi_4_clean_rec >=5 & covid_psqi_4_clean_rec < 6,
             "2",
             comp3_sleep_dur),
    comp3_sleep_dur =
      ifelse(covid_psqi_4_clean_rec < 5,
             "3",
             comp3_sleep_dur)
    ) %>%
  dplyr::select(-covid_psqi_4_copy)
```

### _4_ Habitual sleep efficiency
#### _step 1_ 
Already coded as it should be! 

#### _step 2_
calc the num of hrs spent in bed (getting up time: quest 3) and bedtime (quest #1)    
some individuls recognized military time, and some didn't  
recoding covid_psqi_1 (bedtime) to be all military time  
this mainly affected the following times (7pm - 12pm): mil_time <- 07 = 19, 08 = 20, 09 = 21, 10 = 22, 11 = 23, 12 = 00  
covid_psqi_3 times all made sense  
```{r}
# selecting which obs to recode and adding 12 hrs. (43200 sec), and then extracting just the time
covidt1_comp4_s1 <-
  covidt1_comp3_clean_sum %>%
  mutate(
    covid_psqi_1_rec =
      ifelse(hour(covid_psqi_1) >= 7 & hour(covid_psqi_1) <= 12,
             "recode",
             NA),
    covid_psqi_1_rec_mil =
      ifelse(hour(covid_psqi_1) == 9 | hour(covid_psqi_1) == 10 | hour(covid_psqi_1) == 11 | hour(covid_psqi_1) == 12,
             covid_psqi_1 + 43200,
             covid_psqi_1),
    covid_psqi_1_rec_mil = as_datetime(covid_psqi_1_rec_mil, tz = "UTC")
  )

# unfortunatley, midnight 00:00:00 was not recognized, so recoding that here ("\\.*\\d+\\.*\\d*")
library(stringi)
library(stringr)
covidt1_comp4_s1_recode <- 
  covidt1_comp4_s1 %>%
  mutate(
    covid_psqi_1_rec_mil_hms = str_extract(covid_psqi_1_rec_mil, pattern = "\\d+\\:+\\d+\\:+\\d+"),
  ) %>%
  mutate(
    covid_psqi_1_rec_mil_hm = ifelse(
      hour(covid_psqi_1_rec_mil) == 00 & minute(covid_psqi_1_rec_mil) == 00,
      "00:00:00",
      covid_psqi_1_rec_mil_hms)
  ) %>%
  dplyr::select(-covid_psqi_1_rec_mil_hms, -covid_psqi_1_rec)
```

##### _step 2a_
converting covid_psqi_1_rec_mil_hm to a time object  
converting covid_psqi_1_datetime and covid_psqi_1_datetime to dttm object  
creating "go to bed" date and "getting up time" date  
since some individuals went to bed at/after midnight, had to change those dates to the same date as "get up time" date  
```{r}
library(hms)
covidt1_comp4_s2a <-
  covidt1_comp4_s1_recode %>%
  mutate(
    covid_psqi_1_datetime = as.hms(covid_psqi_1_rec_mil_hm),
    covid_psqi_1_datetime = as_datetime(covid_psqi_1_datetime, tz = "UTC"),
    covid_psqi_3_datetime = as_datetime(covid_psqi_3, tz = "UTC"),
    covid_psqi_1_date = as_date(now()),
    covid_psqi_3_date = covid_psqi_1_date + 1
  ) %>%
  mutate(
    covid_psqi_1_date =
      ifelse(hour(covid_psqi_1_datetime) >= "20" & hour(covid_psqi_1_datetime) <= "24",
             covid_psqi_1_date,
             covid_psqi_3_date)
  ) %>%
  mutate(
    covid_psqi_1_date = as_date(covid_psqi_1_date)
  )
```

##### _step 2b_
updating the covid_psqi_1_datetime and covid_psqi_3_datetime objects with the correct dates
```{r}
covidt1_comp4_s2b <-
  covidt1_comp4_s2a %>%
  mutate(
    covid_psqi_1_datetime = 
      update(covid_psqi_1_datetime, year = year(covid_psqi_1_date), month = month(covid_psqi_1_date), mday = mday(covid_psqi_1_date)),
    covid_psqi_3_datetime =
      update(covid_psqi_3_datetime, year = year(covid_psqi_3_date), month = month(covid_psqi_3_date), mday = mday(covid_psqi_3_date))
  )
```

##### _step 2c_
to get hours in bed subtracting covid_psqi_1_datetime and covid_psqi_3_datetime
```{r}
covidt1_comp4_s2c <-
  covidt1_comp4_s2b %>%
  mutate(
   hrs_in_bed = abs(covid_psqi_1_datetime - covid_psqi_3_datetime)
   ) %>%
  dplyr::select(-covid_psqi_1_rec_mil, -covid_psqi_1_rec_mil_hm, -covid_psqi_1_date, -covid_psqi_3_date)
```


#### _step 3_
calc habitual sleep efficiency (num hrs slept / num hrs in bed) x 100 = habit sleep efficiency (%)  
using the covid_psqi_4_clean_rec for num of hrs slept  
using hrs_in_bed for num of hrs spent in bed  
```{r}
covidt1_comp4_s3 <-
  covidt1_comp4_s2c %>%
  mutate(
    hrs_in_bed = as.numeric(hrs_in_bed),
    sleep_eff = (covid_psqi_4_clean_rec/hrs_in_bed)*100
  )
```

#### _step 4_
assign component 4 score  
```{r}
covidt1_comp4_s4 <-
  covidt1_comp4_s3 %>%
  mutate(
    comp4_sleep_eff =
      ifelse(sleep_eff > 85,
             "0",
             "1"),
    comp4_sleep_eff = 
      ifelse(sleep_eff >= 75 & sleep_eff <= 84,
             "1",
             comp4_sleep_eff),
    comp4_sleep_eff =
      ifelse(sleep_eff >= 65 & sleep_eff <= 74,
             "2",
             comp4_sleep_eff),
    comp4_sleep_eff =
      ifelse(sleep_eff < 65,
             "3",
             comp4_sleep_eff)
    )
```

### _5_ Sleep disturbances
For questions 5b-5j, score like this:    
not during the past month = 0  
less than once a week = 1  
once or twice a week = 3  
three or more times a week = 4  
_step 1 of 5:_ Already coded as it should be!  
#### _step 2_  
sum of 5b-5j
```{r}
covidt1_comp5 <-
  covidt1_comp4_s4 %>%
    mutate(
    sum_sleep_dist = rowSums(across(covid_psqi_5a:covid_psqi_5j))
      )
```

#### _step 3_  
assign component 5 score
```{r}
covidt1_comp5 <-
  covidt1_comp5 %>%
  mutate(
    comp5_sleep_dist =
      ifelse(sum_sleep_dist == 0,
             "0",
             "1"),
    comp5_sleep_dist = 
      ifelse(sum_sleep_dist >= 1 & sum_sleep_dist <= 9,
             "1",
             comp5_sleep_dist),
    comp5_sleep_dist =
      ifelse(sum_sleep_dist >= 10 & sum_sleep_dist <= 18,
             "2",
             comp5_sleep_dist),
    comp5_sleep_dist =
      ifelse(sum_sleep_dist >= 19 & sum_sleep_dist <= 27,
             "3",
             comp5_sleep_dist)
    )
```


### _6_ Use of sleep med  
```{r}
# question 6 in the way that we ordered it ("During the past month, how often did you take medicine
# (prescribed or 'over the counter') to help you sleep?"")
covidt1_comp6 <-
  covidt1_comp5 %>%
  mutate(
    comp6_sleep_med = covid_psqi_6
  )
```

### _7_ Daytime dysfunction
```{r}
# question 7 in the way we ordered it ("During the past month, how often did you have trouble
# staying awake during class time, eating meals, or engaging in social activity?")

# and question 8 in the way we ordered it ("During the past month, how much of a problem was it for
# you to keep up enthusiasm to get things done?")
covidt1_comp7 <-
  covidt1_comp6 %>%
  mutate(
    sum_day_disfunc = rowSums(across(covid_psqi_7:covid_psqi_8))
  )

# assign component 7 score
covidt1_comp7 <-
  covidt1_comp7 %>%
  mutate(
    comp7_day_disfunc =
      ifelse(sum_day_disfunc == 0,
             "0",
             "1"),
    comp7_day_disfunc = 
      ifelse(sum_day_disfunc == 1 | sum_day_disfunc == 2,
             "1",
             comp7_day_disfunc),
    comp7_day_disfunc =
      ifelse(sum_day_disfunc == 3 | sum_day_disfunc == 4,
             "2",
             comp7_day_disfunc),
    comp7_day_disfunc =
      ifelse(sum_day_disfunc == 5 | sum_day_disfunc == 6,
             "3",
             comp7_day_disfunc)
    )
```

_Summing all 7 components together_  
```{r}
to_num <-
  function(x) as.numeric(x)

vars <- c("comp1_subj_sleep_qual", "comp2_sleep_lat", "comp3_sleep_dur", 
    "comp4_sleep_eff", "comp5_sleep_dist", "comp6_sleep_med", "comp7_day_disfunc")
covidt1_comp_sum_all <-
  covidt1_comp7 %>%
  dplyr::select(
    ELS_ID, Child_Age.TC, Hours_sleep_3months.TC, psqi_timestamp, 
    comp1_subj_sleep_qual, comp2_sleep_lat, comp3_sleep_dur, 
    comp4_sleep_eff, comp5_sleep_dist, comp6_sleep_med, comp7_day_disfunc
  ) %>%
  mutate_at(vars, to_num) %>%
  mutate(
    covid_psqi_t1_total = rowSums(across(comp1_subj_sleep_qual:comp7_day_disfunc))
  ) %>%
  drop_na()
```

```{r}
library(cocron)
library(MBESS)
map(covidt1_comp_sum_all, ~sum(is.na(.))) 

psqi_alpha <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc)
  
cronbach.alpha(psqi_alpha, standardized = T)
# 0.56 (NOT GREAT!)
# ci.reliability(data=psqi_alpha, type="hierarchical", conf.level = 0.95,
# interval.type="bca", B=1000)
# varTable(psqi_alpha)
```

#### what happens if I remove a component
```{r}
# remove comp 1
psqi_alpha_rem_c1 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp2_sleep_lat:comp7_day_disfunc)
cronbach.alpha(psqi_alpha_rem_c1, standardized = T) # 0.4554512 (worse!)

# remove comp 2
psqi_alpha_rem_c2 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp2_sleep_lat)
cronbach.alpha(psqi_alpha_rem_c2, standardized = T) # 0.5159377 (worse!)

# remove comp 3
psqi_alpha_rem_c3 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp3_sleep_dur)
cronbach.alpha(psqi_alpha_rem_c3, standardized = T) # 0.5399781 (worse!)

# remove comp 4
psqi_alpha_rem_c4 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp4_sleep_eff)
cronbach.alpha(psqi_alpha_rem_c4, standardized = T) # 0.5925679 (better!)

# remove comp 5
psqi_alpha_rem_c5 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp5_sleep_dist)
cronbach.alpha(psqi_alpha_rem_c5, standardized = T) # 0.4417017 (worse!)

# remove comp 6
psqi_alpha_rem_c6 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp6_sleep_med)
cronbach.alpha(psqi_alpha_rem_c6, standardized = T) # 0.5663596 (better!)

psqi_alpha_rem_c7 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp6_sleep_med)
cronbach.alpha(psqi_alpha_rem_c7, standardized = T) # 0.491033 (worse!)

# removing comps 3, 4 and 6
psqi_alpha_rem_c34and6 <- 
  covidt1_comp_sum_all %>%
  group_by(ELS_ID) %>%
  ungroup() %>%
  dplyr::select(comp1_subj_sleep_qual:comp7_day_disfunc, -comp3_sleep_dur, -comp4_sleep_eff, -comp6_sleep_med)
cronbach.alpha(psqi_alpha_rem_c34and6, standardized = T) # 0.6590607 (better!)


```

_Distribution of global PSQI scores at T1_
```{r}
covid_psqi_hist <-
 covidt1_comp_sum_all %>%
  ggplot(aes(x=covid_psqi_t1_total)) +
  geom_histogram(stat = "count", fill = "purple") +
  theme_minimal() +
  labs(
    x = "PSQI Global Score"
  ) + 
  theme(
    axis.text = element_text(size = 12, angle = 30, hjust = 1),
    axis.title = element_text(size = 14)
  ) 
covid_psqi_hist
```
```{r}
covidt1_comp_sum_all_long <-
  covidt1_comp_sum_all %>%
  gather(component, comp_score, comp1_subj_sleep_qual:comp7_day_disfunc)
```

_Distribution of  PSQI scores by Domain at T1_
```{r}
covid_psqi_hist_bycomp <-
 covidt1_comp_sum_all_long %>%
  ggplot(aes(x=comp_score, fill = component)) +
  geom_histogram(binwidth = .5, position ="dodge") +
  facet_wrap(~component) +
  theme_minimal() +
  labs(
    x = "PSQI Scores",
    title = "Distribution of PSQI Domains"
  ) + 
  theme(
    axis.text = element_text(size = 12, angle = 30, hjust = 1),
    axis.title = element_text(size = 14)
  ) 
covid_psqi_hist_bycomp
ggsave("covid_psqi_hist_bycomp.jpg", covid_psqi_hist_bycomp, width = 7, height = 5)
```

Clearly the domains where most people scored zero (comp 3, 4, and 6) were reducing internal consistency.

After reoming these items
```{r}
covidt1_comp_sum_all <-
  covidt1_comp_sum_all %>%
mutate(
    covid_psqi_t1_total_items1257 = 
      comp1_subj_sleep_qual + comp2_sleep_lat + comp5_sleep_dist + comp7_day_disfunc)
```

```{r}
covid_psqi_hist1257 <-
 covidt1_comp_sum_all %>%
  ggplot(aes(x=covid_psqi_t1_total_items1257)) +
  geom_histogram(stat = "count", fill = "purple") +
  theme_minimal() +
  labs(
    x = "PSQI Global Score (without Domains of Sleep Duration, Efficiency, and Medication"
  ) + 
  theme(
    axis.text = element_text(size = 12, angle = 30, hjust = 1),
    axis.title = element_text(size = 14)
  ) 
covid_psqi_hist1257
```

```{r}
covid_psqi_t1_fp <- "~/JackieSchwartz/COVID19/covid_psqi_t1.csv"
write_csv(covidt1_comp_sum_all, covid_psqi_t1_fp)
```

